---
title: "Weather and Corn Yield Regressions"
author: "Nathan Mueller | Assignment completed by: George Woolsey"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_float: true
    # code_folding: hide
editor_options: 
  chunk_output_type: console
knit: (function(inputFile, encoding) { rmarkdown::render(inputFile, encoding = encoding, output_file = file.path(dirname(inputFile), 'index.html')) })
---

```{r setup, include=FALSE, warning=F, message=F}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lubridate)
library(viridis)
library(RColorBrewer)
library(scales)
####### read data req. for this assignment
library(R.matlab)
library(rnassqs)
####### spatial data
library(USAboundariesData)
library(USAboundaries)
library(sf)
library(mapview) #Interactive maps
library(leafpop) #map html popup
# regression plots
library(ggiraph)
library(ggiraphExtra)


#define vars
my_qs_api_key <- "1E804485-6B71-3E4A-B74E-2340256BC471"
filter <- dplyr::filter

```

## Weather Data Analysis

### Load the PRISM daily maximum temperatures

```{r, warning=F, message=F, results='hide', fig.width = 8, fig.height = 5}

# daily max temperature
# dimensions: counties x days x years
prism <- readMat("data/prismiowa.mat")

# look at county #1
t_1981_c1 <- prism$tmaxdaily.iowa[1,,1]
t_1981_c1[366]
plot(1:366, t_1981_c1, type = "l")

ggplot() +
  geom_line(mapping = aes(x=1:366, y = t_1981_c1)) +
  theme_bw() +
  xlab("day of year") +
  ylab("daily maximum temperature (째C)") +
  ggtitle("Daily Maximum Temperature, Iowa County #1")


```

### tidy weather data

```{r, warning=F, message=F, results='hide', fig.width = 8, fig.height = 5}

# assign dimension names to tmax matrix
dimnames(prism$tmaxdaily.iowa) <- list(prism$COUNTYFP, 1:366, prism$years)

# converted 3d matrix into a data frame
tmaxdf <- as.data.frame.table(prism$tmaxdaily.iowa)

# relabel the columns
colnames(tmaxdf) <- c("countyfp","doy","year","tmax")
tmaxdf <- tibble(tmaxdf)

```

## Temperature trends

### Summer temperature trends: Winneshiek County

```{r, warning=F, message=F, results='hide', fig.width = 8, fig.height = 5}

tmaxdf$doy <- as.numeric(tmaxdf$doy)
tmaxdf$year <- as.numeric(as.character(tmaxdf$year))

winnesummer <- tmaxdf %>%
  filter(countyfp==191 & doy >= 152 & doy <= 243) %>%
  group_by(year) %>%
  summarize(meantmax = mean(tmax))

ggplot(winnesummer, mapping = aes(x = year, y = meantmax)) +
  geom_point() +
  theme_bw() +
  labs(x = "year", y = "Tmax (째C)") +
  geom_smooth(method = lm)

lm_summertmax <- lm(meantmax ~ year, winnesummer)
summary(lm_summertmax)

```

### Winter Temperatures - Winneshiek County

```{r, warning=F, message=F, results='hide', fig.width = 8, fig.height = 5}

winnewinter <- tmaxdf %>%
  filter(countyfp==191 & (doy <= 59 | doy >= 335) & !is.na(tmax)) %>%
  group_by(year) %>%
  summarize(meantmax = mean(tmax))

ggplot(winnewinter, mapping = aes(x = year, y = meantmax)) +
  geom_point() +
  theme_bw() +
  labs(x = "year", y = "Tmax (째C)") +
  geom_smooth(method = lm)

lm_wintertmax <- lm(meantmax ~ year, winnewinter)
summary(lm_wintertmax)

```

### Multiple regression -- Quadratic time trend

```{r, warning=F, message=F, results='hide', fig.width = 8, fig.height = 5}

winnewinter$yearsq <- winnewinter$year^2

lm_wintertmaxquad <- lm(meantmax ~ year + yearsq, winnewinter)
summary(lm_wintertmaxquad)
winnewinter$fitted <- lm_wintertmaxquad$fitted.values

ggplot(winnewinter) +
  geom_point(mapping = aes(x = year, y = meantmax)) +
  geom_line(mapping = aes(x = year, y = fitted)) +
  theme_bw() +
  labs(x = "year", y = "tmax")

```

### Download NASS corn yield data

```{r, warning=F, message=F, results='hide', fig.width = 8, fig.height = 5}

# set our API key with NASS
nassqs_auth(key = my_qs_api_key)

# parameters to query on 
params <- list(commodity_desc = "CORN", util_practice_desc = "GRAIN", prodn_practice_desc = "ALL PRODUCTION PRACTICES", year__GE = 1981, state_alpha = "IA")

# download
cornyieldsall <- nassqs_yields(params)

cornyieldsall$county_ansi <- as.numeric(cornyieldsall$county_ansi)
cornyieldsall$yield <- as.numeric(cornyieldsall$Value)

# clean and filter this dataset
cornyields <- select(cornyieldsall, county_ansi, county_name, yield, year) %>%
  filter(!is.na(county_ansi) & !is.na(yield))
cornyields <- tibble(cornyields) %>% 
  rename(year_id = year
         , yield_corn_bu_acre = yield # bushels per acre
    )

```

## Assignment

### Question 1a
Extract Winneshiek County corn yields, fit a linear time trend, make a plot. Is there a significant time trend?

```{r, warning=F, message=F, results='hide', fig.width = 8, fig.height = 5}
# # check out yield data structure
# str(cornyields)
# nrow(cornyields)
# cornyields %>% distinct(county_ansi) %>% nrow(.)
# cornyields %>% distinct(county_ansi, year_id) %>% nrow(.)
# # row is unique by county, year

# filter to specific county
winneshiek_yields <- cornyields %>% 
  filter(toupper(county_name) == "WINNESHIEK")

# trend model for yield growth
lm_yield_year <- lm(yield_corn_bu_acre ~ year_id , data = winneshiek_yields)

summary(lm_yield_year)$coefficients["(Intercept)", "Estimate"]

# label model in ggplot
model_label <- as.character(as.expression(
  substitute(
    italic(y) == a + b %.% italic(year)*","~~italic(r)^2~"="~r2
    , list(
        a = scales::comma(summary(lm_yield_year)$coefficients["(Intercept)", "Estimate"], accuracy = 0.1)
        , b = scales::comma(summary(lm_yield_year)$coefficients[2, "Estimate"], accuracy = 0.1)
        , r2 = format(summary(lm_yield_year)$r.squared, digits = 2)
      )
  )
))
# model_label

# plot
ggplot(winneshiek_yields, aes(x = year_id, y = yield_corn_bu_acre)) +
  geom_point(alpha=0.8, color = "navy", size = 2) + 
  geom_smooth(method = 'lm', color = "grey35") +
  annotate(geom = "text", label = model_label, parse = TRUE, x = -Inf, y = Inf, hjust = -0.1, vjust = 2) +
  scale_x_continuous(breaks = scales::extended_breaks(n=10)) +
  labs(
    title = "Corn Yield over time in  Winneshiek County, Iowa"
  ) +
  xlab("Year") +
  ylab("Corn Yield (bushels per acre)") +
  theme_bw() +
  theme(
    legend.position="none"
  )
```

<span style="color: blue;">There is a significant time trend in the corn yields in  Winneshiek County, Iowa from `r as.character(min(winneshiek_yields$year_id))` to `r as.character(max(winneshiek_yields$year_id))`. Based on a simple linear regression, every additional year results in the corn yield increasing by **`r round(summary(lm_yield_year)$coefficients[2, "Estimate"], 2)` bu/acre**. The annual time trend explains **`r scales::percent(summary(lm_yield_year)$r.squared)`** of the variation in corn yields in  Winneshiek County, Iowa.</span>

### Question 1b
Fit a quadratic time trend (i.e., year + year^2) and make a plot. Is there evidence for slowing yield growth? 

```{r, warning=F, message=F, fig.width = 8, fig.height = 5}
# quadratic trend model for evidence of slowing yield growth
winneshiek_yields <- winneshiek_yields %>% 
  mutate(year_id_sq = year_id^2)

# run model
lm_yield_year_quad <- lm(yield_corn_bu_acre ~ year_id + year_id_sq, data = winneshiek_yields)
# summary(lm_yield_year_quad)
# save fitted values
winneshiek_yields$fitted_quad <- lm_yield_year_quad$fitted.values

# compare models
anova_m1_m2 <- anova(lm_yield_year, lm_yield_year_quad)
# anova_m1_m2[2, "Pr(>F)"]

# label model in ggplot
model_label_quad <- as.character(as.expression(
  substitute(
    italic(y) == a + b %.% italic(year) + c %.% ~italic(year)^2~","~~italic(r)^2~"="~r2
    , list(
        a = scales::comma(summary(lm_yield_year_quad)$coefficients["(Intercept)", "Estimate"], accuracy = 0.1)
        , b = scales::comma(summary(lm_yield_year_quad)$coefficients[2, "Estimate"], accuracy = 0.2)
        , c = scales::comma(summary(lm_yield_year_quad)$coefficients[3, "Estimate"], accuracy = 0.0001)
        , r2 = format(summary(lm_yield_year_quad)$r.squared, digits = 2)
      )
  )
))
# model_label_quad

# plot
ggplot(winneshiek_yields, aes(x = year_id, y = yield_corn_bu_acre)) +
  geom_point(alpha=0.8, color = "navy", size = 2) +
  geom_line(aes(x = year_id, y = fitted_quad), color = "grey35") +
  annotate(geom = "text", label = model_label_quad, parse = TRUE, x = -Inf, y = Inf, hjust = -0.1, vjust = 2) +
  scale_x_continuous(breaks = scales::extended_breaks(n=10)) +
  labs(
    title = "Corn Yield over time in  Winneshiek County, Iowa"
    , subtitle = "with fitted values from quadratic time trend (i.e., year + year^2)"
  ) +
  xlab("Year") +
  ylab("Corn Yield (bushels per acre)") +
  theme_bw() +
  theme(
    legend.position="none"
  )

```

<span style="color: blue;">Using a quadratic time trend (i.e., year + year^2) shows a significant time trend in the corn yields in  Winneshiek County, Iowa from `r as.character(min(winneshiek_yields$year_id))` to `r as.character(max(winneshiek_yields$year_id))`. Based on a the quadratic model, there is a slightly *increasing* rate of corn yield over time. The quadratic time trend explains **`r scales::percent(summary(lm_yield_year_quad)$r.squared)`** of the variation in corn yields in  Winneshiek County, Iowa. Using the `anova()` function to compare the quadratic model with the simple linear model allows for us to test which model provides the best parsimonious fit of the data. Although the quadratic model is significant, the ANOVA test has a resulting p-value of **`r round(anova_m1_m2[2, "Pr(>F)"], 3)`**. This result means that adding the quadratic term did not significantly improve the model over the simple linear model.</span>


### Question 2
Time Series: Let's analyze the relationship between temperature and yields for the Winneshiek County time series. Use data on yield and summer avg Tmax. Is adding year or Tmax^2 to your model helpful? Make a plot and interpret the results.

#### Yields on Temperature

```{r, warning=F, message=F, fig.width = 8, fig.height = 5}
#join summer temperature to yield data for Winneshiek County
winneshiek_yields_temp <- inner_join(winneshiek_yields, winnesummer, by = c("year_id"= "year"))

# yield on temperature
lm_yield_tmax <- lm(yield_corn_bu_acre ~ meantmax, data = winneshiek_yields_temp)
lm_yield_yr <- lm(yield_corn_bu_acre ~ year_id, data = winneshiek_yields_temp)
# summary(lm_yield_tmax)

# compare models
anova_m1_m3 <- anova(lm_yield_yr, lm_yield_tmax)

## plot
ggplot(winneshiek_yields_temp, aes(x = meantmax, y = yield_corn_bu_acre)) +
  geom_point(alpha=0.8, color = "navy", size = 2) + 
  geom_smooth(method = 'lm', color = "grey35") +
  scale_x_continuous(breaks = scales::extended_breaks(n=10)) +
  labs(
    title = "Corn Yield versus Mean Summer Max. Temperature in  Winneshiek County, Iowa"
  ) +
  xlab("Summer Max. Temp. (째C)") +
  ylab("Corn Yield (bushels per acre)") +
  theme_bw() +
  theme(
    legend.position="none"
  )
```

<span style="color: blue;">A simple linear regression of corn yield on mean summer maximum temperature shows that temperature alone explains **`r scales::percent(summary(lm_yield_tmax)$r.squared)`** of the variation in corn yields in  Winneshiek County, Iowa. This model performs worse than the simple time trend of annual corn yield investigated above.</span>

#### Yields on Year & Temperature

```{r, warning=F, message=F, fig.width = 8, fig.height = 5}
# yield on temperature and year
lm_yield_year_tmax <- lm(yield_corn_bu_acre ~ year_id + meantmax, data = winneshiek_yields_temp)
# summary(lm_yield_year_tmax)

# compare models
anova_m1_m4 <- anova(lm_yield_yr, lm_yield_year_tmax)
anova_m1_m4

# linear function holding tmax constant at mean
lm_yield_year_tmax_eq1 <- function(x){summary(lm_yield_year_tmax)$coefficients["(Intercept)", "Estimate"] + (summary(lm_yield_year_tmax)$coefficients[2, "Estimate"] * x) + (summary(lm_yield_year_tmax)$coefficients[3, "Estimate"] * mean(winneshiek_yields_temp$meantmax))}

#label
model_label_yr_tmax <- as.character(as.expression(
  substitute(
    italic(y) == a + b %.% italic(year) + c %.% ~italic(tmax)~","~~italic(r)^2~"="~r2
    , list(
        a = scales::comma(summary(lm_yield_year_tmax)$coefficients["(Intercept)", "Estimate"], accuracy = 1)
        , b = scales::comma(summary(lm_yield_year_tmax)$coefficients[2, "Estimate"], accuracy = 0.2)
        , c = scales::comma(summary(lm_yield_year_tmax)$coefficients[3, "Estimate"], accuracy = 0.2)
        , r2 = format(summary(lm_yield_year_tmax)$r.squared, digits = 2)
      )
  )
))

#plot
ggplot(winneshiek_yields_temp, aes(x = year_id, y = yield_corn_bu_acre)) +
  geom_point(alpha=0.8, color = "navy", size = 2) +
  stat_function(fun = lm_yield_year_tmax_eq1
      , geom="line"
      , show.legend = FALSE
      , color = "grey35"
    ) +
  annotate(geom = "text", label = model_label_yr_tmax, parse = TRUE, x = -Inf, y = Inf, hjust = -0.1, vjust = 2) +
  scale_x_continuous(breaks = scales::extended_breaks(n=10)) +
  labs(
    title = "Corn Yield over time in Winneshiek County, Iowa"
    , subtitle = "with fitted values of yield over time while holding max. temperature constat at mean"
  ) +
  xlab("Year") +
  ylab("Corn Yield (bushels per acre)") +
  theme_bw() +
  theme(
    legend.position="none"
  )

```

<span style="color: blue;">A multiple linear regression of corn yield on year and mean summer maximum temperature explains **`r scales::percent(summary(lm_yield_year_tmax)$r.squared)`** of the variation in corn yields in  Winneshiek County, Iowa. Using the `anova()` function to compare the model including year and temperature with the simple linear model including only year allows for us to test which model provides the best parsimonious fit of the data. The ANOVA test has a resulting p-value of **`r round(anova_m1_m4[2, "Pr(>F)"], 3)`**. This result means that adding the mean summer maximum temperature did not significantly improve the model over the simple linear model on year alone.</span>

#### Yields on Year & Temperature quadratic

```{r, warning=F, message=F, fig.width = 8, fig.height = 5}
# yield on temperature quadratic and year
winneshiek_yields_temp <- winneshiek_yields_temp %>% 
  mutate(meantmax_sq = meantmax^2)

lm_yield_year_tmax_sq <- lm(yield_corn_bu_acre ~ year_id + meantmax + meantmax_sq, data = winneshiek_yields_temp)
# summary(lm_yield_year_tmax_sq)

# compare models
anova_m1_m5 <- anova(lm_yield_yr, lm_yield_year_tmax_sq)
anova_m1_m5

# linear function holding tmax constant at mean
lm_yield_year_tmax_sq_eq1 <- function(x){summary(lm_yield_year_tmax_sq)$coefficients["(Intercept)", "Estimate"] + (summary(lm_yield_year_tmax_sq)$coefficients[2, "Estimate"] * x) + (summary(lm_yield_year_tmax_sq)$coefficients[3, "Estimate"] * mean(winneshiek_yields_temp$meantmax))  + (summary(lm_yield_year_tmax_sq)$coefficients[4, "Estimate"] * mean(winneshiek_yields_temp$meantmax)^2)}

#label
model_label_yr_tmax_sq <- as.character(as.expression(
  substitute(
    italic(y) == a + b %.% italic(year) + c %.% ~italic(tmax) + d %.% ~italic(tmax)^2~","~~italic(r)^2~"="~r2
    , list(
        a = scales::comma(summary(lm_yield_year_tmax_sq)$coefficients["(Intercept)", "Estimate"], accuracy = 1)
        , b = scales::comma(summary(lm_yield_year_tmax_sq)$coefficients[2, "Estimate"], accuracy = 0.2)
        , c = scales::comma(summary(lm_yield_year_tmax_sq)$coefficients[3, "Estimate"], accuracy = 0.2)
        , d = scales::comma(summary(lm_yield_year_tmax_sq)$coefficients[4, "Estimate"], accuracy = 0.2)
        , r2 = format(summary(lm_yield_year_tmax_sq)$r.squared, digits = 2)
      )
  )
))

#plot
ggplot(winneshiek_yields_temp, aes(x = year_id, y = yield_corn_bu_acre)) +
  geom_point(alpha=0.8, color = "navy", size = 2) +
  stat_function(fun = lm_yield_year_tmax_sq_eq1
      , geom="line"
      , show.legend = FALSE
      , color = "grey35"
    ) +
  annotate(geom = "text", label = model_label_yr_tmax_sq, parse = TRUE, x = -Inf, y = Inf, hjust = -0.02, vjust = 2) +
  scale_x_continuous(breaks = scales::extended_breaks(n=10)) +
  labs(
    title = "Corn Yield over time in Winneshiek County, Iowa"
    , subtitle = "with fitted values of yield over time while holding max. temperature constat at mean"
  ) +
  xlab("Year") +
  ylab("Corn Yield (bushels per acre)") +
  theme_bw() +
  theme(
    legend.position="none"
  )

```

<span style="color: blue;">A multiple linear regression of corn yield on year and mean summer maximum temperature explains **`r scales::percent(summary(lm_yield_year_tmax_sq)$r.squared)`** of the variation in corn yields in  Winneshiek County, Iowa. Using the `anova()` function to compare the model including year and temperature with the simple linear model including only year allows for us to test which model provides the best parsimonious fit of the data. The ANOVA test has a resulting p-value of **`r as.character(round(anova_m1_m5[2, "Pr(>F)"], 5))`**. This result means that adding the mean summer maximum temperature with a squared term did significantly improve the model over the simple linear model on year alone. Thus, of all of the models tested to estimate corn yields in  Winneshiek County, Iowa, this model provides the best parsimonious fit of the data.</span>

### Question 3
Cross-Section: Analyze the relationship between temperature and yield across all counties in 2018. Is there a relationship? Interpret the results.

```{r, warning=F, message=F, results='hide', fig.width = 8, fig.height = 5}
#code

```


### Question 4
Panel: One way to leverage multiple time series is to group all data into what is called a "panel" regression. Convert the county ID code ("countyfp" or "county_ansi") into factor using as.factor, then include this variable in a regression using all counties' yield and summer temperature data. How does the significance of your temperature coefficients (Tmax, Tmax^2) change? Make a plot comparing actual and fitted yields and interpret the results of your model.

```{r, warning=F, message=F, results='hide', fig.width = 8, fig.height = 5}
#code

```


### Question 5
Soybeans: Download NASS data on soybean yields and explore either a time series relationship for a given county, the cross-sectional relationship for a given year, or a panel across all counties and years.

```{r, warning=F, message=F, results='hide', fig.width = 8, fig.height = 5}
#code

```


### Bonus
Find a package to make a county map of Iowa displaying some sort of information about yields or weather. Interpret your map.

```{r, warning=F, message=F, results='hide', fig.width = 8, fig.height = 5}
# pull in spatial county data to join on count FIPS code
counties <- us_counties() %>%
  setNames(make.names(names(.), unique = TRUE)) %>% # for some reason the state_name field is duplicated
  mutate(county_ansi = as.numeric(countyfp)) %>%
  filter(state_name == "Iowa") %>% 
  st_transform(2163) # EPSG:2163 = US National Atlas Equal Area

# mapview(counties)

# join to yields
spatial_county_yields <- inner_join(counties, cornyields, by = "county_ansi")
# mapview(spatial_county_yields %>% filter(year_id == 2001), zcol = "yield_corn_bu_acre")


```


### Bonus #2
Challenge question - map trends in corn yields by county across Iowa. Interpret your map.

```{r, warning=F, message=F, results='hide', fig.width = 8, fig.height = 5}
#code

```

